Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-27 manuscript/paper.tex
polish intro

\section{Introduction} 
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a recombining
species, and are currently the subject of intense research 
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
gunnarsson2024scalable,deng2024robust} that practical application has 
become possible. Varied applications
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and 
the spatial 
location of genetic
ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
more powerful approaches to quantifying genetic 
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association 
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree} all depend critically on the \emph{accuracy}
of the inferred ARGs. 
Although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data is 
understudied, and there is a critial lack of software infrastructure
to support evaluating the quality of estimated ARGs.

Visualisation is fundamentally important to data analysis.
Many specialised tools exists to aid the visual analysis and 
quality control
of assembly~\citep{wick2015bandage,challis2020blobtoolkit},
read mapping~\citep{robinson2011integrative},
and variant
calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
for example.
At every stage of a bioinformatics pipeline, it is important to visualise 
results to avoid artefacts and aid understanding of the data.
Genome browsers such as IGV~\citep{robinson2011integrative} and the 
UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
data modalities, and are vital infrastructure for the field.
% . In the past few decades,
% genome browsers have become synonymous with sequence visualisation, by
% providing user-friendly, interactive interfaces to navigate genomic sequences
% and annotations. From early standalone tools to manage \textit{C. elegans} data
% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
% with the changing landscape of genomics. Modern genome browsers can load
% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
% The data security and scalability offered by this architecture ensures that
% genome browsers continue to be relevant in the biobank era.

There is currently no straightforward means of visually summarising
ARGs, presenting a significant stumbling block for the nascent 
field of practical ARG inference. Inferred ARGs are essentially opaque,
with only the most basic numerical summaries (such as numbers 
of nodes, mutations etc) or high-level
statistics~\citep{ralph2020efficiently} available. While tools 
for visualising the local tree topologies exist, they are difficult
to interpret and do not scale well to large sample sizes. 
To address this gap, we present \texttt{tsbrowse}, a
% I worry that if we say "genome browser" we'll get stung for not having 
% deep integration with all sorts of stuff
client-server application providing genome browser-like 
functionality for ARGs. 
It provides interactive visualisations of the information structure 
of ARGs, smoothly scrolling from chromosome-level views down to 
individual nodes, edges and mutations. Supporting very large ARGs is a 
particular focus for \texttt{tsbrowse}, as millions of genome
sequences have been sampled for several
species~\citep{cesarani2022multibreed,stark2024call,hunt2024addressing}
and ARGs of this scale being actively
inferred~\citep{kelleher2019inferring,zhang2023biobank,
zhan2023towards,gunnarsson2024scalable}.

>>>>

\section{Introduction} 
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a recombining
species, and are currently the subject of intense research 
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
gunnarsson2024scalable,deng2024robust} that practical application has 
become possible. Varied applications have been recently proposed,
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and 
the spatial 
location of genetic
ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
more powerful approaches to quantifying genetic 
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association 
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree}.
While these developements are exciting, the performance 
of these new methods depend critically on the accuracy of the inferred ARGs,
and although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data is 
understudied. In particular, there is a critial lack of software 
infrastructure to support evaluation and quality control of 
inferred ARGs.

Visualisation is fundamentally important to data analysis.
Many specialised tools exist to aid the visual analysis and 
quality control
of assembly~\citep{wick2015bandage,challis2020blobtoolkit},
read mapping~\citep{robinson2011integrative},
and variant
calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
for example.
At every stage of a bioinformatics pipeline, it is important to visualise 
results to avoid artefacts and aid understanding of the data.
Genome browsers such as IGV~\citep{robinson2011integrative} and the 
UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
data modalities, and are vital infrastructure for the field.
% . In the past few decades,
% genome browsers have become synonymous with sequence visualisation, by
% providing user-friendly, interactive interfaces to navigate genomic sequences
% and annotations. From early standalone tools to manage \textit{C. elegans} data
% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
% with the changing landscape of genomics. Modern genome browsers can load
% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
% The data security and scalability offered by this architecture ensures that
% genome browsers continue to be relevant in the biobank era.

There is currently no straightforward means of visually summarising
ARGs, presenting a significant stumbling block for the nascent 
field of practical ARG inference. Inferred ARGs are essentially opaque,
with only the most basic numerical summaries (such as numbers 
of nodes, mutations etc) or high-level
statistics~\citep{ralph2020efficiently} available. While tools 
for visualising the local tree topologies exist, they are difficult
to interpret and do not scale well to large sample sizes. 
To address this gap, we present \texttt{tsbrowse}, a
% I worry that if we say "genome browser" we'll get stung for not having 
% deep integration with all sorts of stuff
client-server application providing genome browser-like 
functionality for ARGs. 
It provides interactive visualisations of the information structure 
of ARGs, smoothly scrolling from chromosome-level views down to 
individual nodes, edges and mutations. Supporting very large ARGs is a 
particular focus for \texttt{tsbrowse}, as millions of genome
sequences have been sampled for several
species~\citep{cesarani2022multibreed,stark2024call,hunt2024addressing}
and ARGs of this scale are a particular focus of ongoing 
research~\citep{kelleher2019inferring,zhang2023biobank,
zhan2023towards,anderson2023on,gunnarsson2024scalable}.
